A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-11-14 analysis/paper/paper.Rmd
Added a more detailed background map for the study area

# Point features with city names
cpts <- st_read(here('analysis/data/raw_data/central_points.gpkg'))

@@

bound_reproj <- st_transform(bboxsitespoly, st_crs(bmap))
bmap2 <- bmap %>%
    filter(st_intersects(., bound_reproj, sparse = FALSE))
bmap_reproj <- st_transform(bmap2, st_crs(sites_sa))

anc <- as.numeric(c(bboxsites$ymin, bboxsites$xmax))
site_map <- ggplot() +
  geom_sf(data = bmap_reproj, fill = "grey", colour = NA) +
  geom_sf(data = isobases, aes(colour = name)) +
  geom_sf(data = st_centroid(sites_sa), fill = "red",
          size = 2.25, shape = 21,
          colour = "black") +
  ggsn::scalebar(data = sites_sa, dist = 20, dist_unit = "km",
                 transform = FALSE, st.size = 4, height = 0.02,
                 border.size = 0.1, st.dist = 0.03,
                 anchor = c(x = anc[2] - 15500, y = anc[1]) + 8000) +
  coord_sf(xlim = c(bboxsites[1], bboxsites[3]),
           ylim = c(bboxsites[2], bboxsites[4]),
           expand = FALSE) +
  scale_colour_manual(values = c("Arendal" = "black",
                                   "Larvik" = "darkgreen",
                                   "Tvedestrand" = "blue",
                                   "Horten" = "darkorange")) +
  theme_bw() + theme(axis.title=element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_blank(),
                     rect = element_rect(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_blank(),
                     legend.position = "none")

# Rename to use the precise placenames
rname_disp <- displacement_curves
rname_disp$name <- ifelse(displacement_curves$name == "Larvik",
                                   "Gunnarsrød", 
                                   ifelse(
                                     displacement_curves$name == "Tvedestrand",
                                     "Hanto", 
                                     ifelse(
                                       displacement_curves$name == "Arendal",
                                       "Bjørnebu", "Skoppum")))

@@

# Plot of dates
date_plot <- plot_dates(datedat, sitename, title = FALSE)
site_map <- site_plot(sitearea, sitel, 150, date_groups,
                      s_tdist = 0.5, s_xpos = 175,
                       s_ypos = 85,  s_bheight = 0.3, s_tsize = 3)
overview_map <- overview_plot(bmap, sitel, sites_sa, isobases,
                               os_tdist = 0.03, os_xpos = 20000, os_ypos = 8000,
                               os_bheight = 0.02, os_tsize = 3)

curve_plot <- ggplot2::ggplot() + 
              geom_line(data = rname_disp, 
                      aes(x = years, y = upperelev, col = name), alpha = 0.4) +
  geom_line(data =  rname_disp, 
                      aes(x = years, y = lowerelev, col = name), alpha = 0.4) +
  geom_line(data = sitecurve,
            aes(x = years, y = upperelev, col = name)) +
  geom_line(data = sitecurve,
            aes(x = years, y = lowerelev, col = name)) +
  scale_colour_manual(values = c("Skoppum" = "darkorange",
                                 "Gunnarsrød" = "darkgreen",
                                 "Hanto" = "blue",
                                 "Bjørnebu" = "black",
                                 "Hegna vest 1" = "red")) +
  ylab("Meters above present sea-level") +
  xlab("BCE/CE") +
  theme_bw() +
  theme(legend.title = element_blank(), legend.position = "bottom",
        legend.direction = "horizontal")

@@

The negative values around 8000 BCE originate from the sites Løvås 1, 2 and 3. @berg-hansen2022[644] made a similar observation in their assessment of the correspondence between shoreline displacement and radiocarbon dates from these sites. These are recently excavated, well-dated sites situated in a relatively undisturbed area of the landscape [@reitan2022b; @berg-hansen2022]. While there could be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and, in turn, use these to evaluate the degree of shore-bound settlement, the sites do clearly represent  
an upper constraining limit for the sea-level, as they would not have been in use when located under water. It therefore seems that the Løvås sites represent a case where the archaeological material indicates a slight discrepancy in the geologic reconstruction of shoreline displacement in the area.

@@

It is clear that the distributions in Figure \@ref(fig:results2) have a severe right skew. Most sites were likely situated less than a meter from the shoreline, and from this there is a sharp decline in density as one moves further along the x-axes. To characterise this relationship, a series of standard models for distributions with a right skew have been fit to the simulation results for vertical distance older than 2500 BCE (Figure \@ref(fig:results2)A) by means of maximum likelihood estimation (Table \@ref(tab:tab1)). As most of the models only accept positive values a constant of 0.001 was added to avoid values of zero. It was attempted to both remove negative values and force these to zero before adding the constant. As the difference between these two solutions was negligible, and as the assumption here is that negative values in actuality reflect a distance of zero, the latter approach was chosen (a plot displaying the negative values and the compared models is available in the supplementary material).

@@

An example of an implementation of the outlined approach is given in Figure \@ref(fig:exampledat), where $\alpha$ = `r round(gammafit$estimate[1], 3)` and $\sigma$ = `r round(gammafit$estimate[2], 3)`. These are the parameters for the gamma distribution identified when considering all pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A) and are the parameters used in all applications of the proposed method that follow below. For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, which, following from the adjustment of the values for fitting the gamma distribution, starts from 0.001m. The gamma distribution is stepped through in it's cumulative form, where the probability from the previous 0.001m step is subtracted from the probability at the current step. This probability is then divided equally across the individual calendar years in the range between the lower and the upper limit of the displacement curve at the current 0.001m step. The probability mass function that is the resulting shoreline date is the sum of performing this procedure on all possible 0.001m values of $D$, which, in practice, is down to and including $E - D = 0.001$ or when 99.999% of the gamma distribution has been stepped through.

@@

```{r exampledat, echo = FALSE, fig.cap = paste0("Shoreline dating of Hegna vest 1. The mean elevation of the site polygon is used to inform $E$ in the dating of the site. The gamma distribution in blue on the y-axis extends the full range of possible values for $E - D$ and has the parameters $\\alpha$ = ", round(gammafit$estimate[1], 3), " and $\\sigma$ = ", round(gammafit$estimate[2], 3),  " (see Figure 7A). The red line marks the shoreline displacement curve interpolated to the site location. The resulting shoreline date in grey is underlined with the 95\\% HDR in black."), out.width = "500px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/sdate.png"))

@@

With these considerations in mind, the results also indicate that shoreline dating has at times been applied with an exaggerated degree of precision. While the implications of a more stable RSL-change for shoreline dating are well known, this also appears to be somewhat under-appreciated in the practical implementation of the method. The results indicate that the spatial and temporal contingency of the method is better captured by the implementation suggested here, as is illustrated by the variation in the range of the 95% HDRs for the dates. In some cases the proposed method provides a relatively precise date and in others it offers little more than a *terminus post quem*. This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise dates (cf. Figure \@ref(fig:map-iso)B). However, as some of the 95% HDRs extend well beyond major chronological divisions, even into the Iron Age, it is also clear that some of these could be severely and securely constrained with only cursory reference to typology. While this would be trivial in some cases, the nature and uncertainty inherent to the method still means that this is arguably a required exercise that should be explicitly performed. This also points to the possibility of drawing on other temporal data to further improve the precision of the dates that can be achieved with shoreline dating.

>>>>

# Background map, more detailed for Norway
bmap2 <- st_read(here('analysis/data/raw_data/naturalearth_norway.gpkg'))

# Point features with city names
cpts <- st_read(here('analysis/data/raw_data/naturalearth_cities.gpkg'))

@@

bound_reproj <- st_transform(bboxsitespoly, st_crs(sites_sa))
bmap2 <- bmap2 %>%
    filter(st_intersects(., bound_reproj, sparse = FALSE))
bmap2_reproj <- st_transform(bmap2, st_crs(sites_sa))

anc <- as.numeric(c(bboxsites$ymin, bboxsites$xmax))
site_map <- ggplot() +
  geom_sf(data = bmap2_reproj, fill = "grey", colour = NA) +
  geom_sf(data = isobases, aes(colour = name)) +
  geom_sf(data = st_centroid(sites_sa), fill = "red",
          size = 2.25, shape = 21,
          colour = "black") +
  ggsn::scalebar(data = sites_sa, dist = 20, dist_unit = "km",
                 transform = FALSE, st.size = 4, height = 0.02,
                 border.size = 0.1, st.dist = 0.03,
                 anchor = c(x = anc[2] - 15500, y = anc[1]) + 8000) +
  coord_sf(xlim = c(bboxsites[1], bboxsites[3]),
           ylim = c(bboxsites[2], bboxsites[4]),
           expand = FALSE) +
  scale_colour_manual(values = c("Arendal" = "black",
                                   "Larvik" = "darkgreen",
                                   "Tvedestrand" = "blue",
                                   "Horten" = "darkorange")) +
  theme_bw() + theme(axis.title=element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_blank(),
                     rect = element_rect(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_blank(),
                     legend.position = "none")

# Rename to use the precise placenames
rname_disp <- displacement_curves
rname_disp$name <- ifelse(displacement_curves$name == "Larvik",
                                   "Gunnarsrød", 
                                   ifelse(
                                     displacement_curves$name == "Tvedestrand",
                                     "Hanto", 
                                     ifelse(
                                       displacement_curves$name == "Arendal",
                                       "Bjørnebu", "Skoppum")))
rname_disp$name <- factor(as.factor(rname_disp$name),
                          levels = c("Bjørnebu", "Hanto",
                                     "Gunnarsrød", "Skoppum"))

@@

dispdat <- rbind(rname_disp, sitecurve)

# Plot of dates
date_plot <- plot_dates(datedat, sitename, title = FALSE)
site_map <- site_plot(sitearea, sitel, 150, date_groups,
                      s_tdist = 0.5, s_xpos = 175,
                       s_ypos = 85,  s_bheight = 0.3, s_tsize = 3)
overview_map <- overview_plot(bmap, sitel, sites_sa, isobases,
                               os_tdist = 0.03, os_xpos = 20000, os_ypos = 8000,
                               os_bheight = 0.02, os_tsize = 3,
                              detailed_map = bmap2)

curve_plot <- ggplot2::ggplot() + 
              geom_line(data = dispdat, 
                      aes(x = years, y = upperelev, col = name), alpha = 0.4) +
  geom_line(data =  dispdat, 
                      aes(x = years, y = lowerelev, col = name), alpha = 0.4) +
  geom_line(data = sitecurve,
            aes(x = years, y = upperelev, col = name)) +
  geom_line(data = sitecurve,
            aes(x = years, y = lowerelev, col = name)) +
  scale_colour_manual(values = c("Skoppum" = "darkorange",
                                 "Gunnarsrød" = "darkgreen",
                                 "Hanto" = "blue",
                                 "Bjørnebu" = "black",
                                 "Hegna vest 1" = "red")) +
  ylab("Meters above present sea-level") +
  xlab("BCE/CE") +
  theme_bw() +
  theme(legend.title = element_blank(), legend.position = "bottom",
        legend.direction = "horizontal")

@@

The negative values around 8000 BCE originate from the sites Løvås 1, 2 and 3. @berg-hansen2022[644] made a similar observation in their assessment of the correspondence between shoreline displacement and radiocarbon dates from these sites. The sites are recently excavated, well-dated and situated in a relatively undisturbed area of the landscape [@reitan2022b; @berg-hansen2022]. While there could be a danger of circularity of having archaeological sites inform a reconstruction RSL-change, and, in turn, use these to evaluate the degree of shore-bound settlement, the sites do clearly represent  
an upper constraining limit for the sea-level, as they would not have been in use when located under water. It therefore seems that the Løvås sites represent a case where the archaeological material indicates a slight discrepancy in the geologic reconstruction of shoreline displacement in the area.

@@

It is clear that the distributions in Figure \@ref(fig:results2) have a severe right skew. Most sites were likely situated less than a meter from the shoreline, and from this there is a sharp decline in density as one moves further along the x-axes. To characterise this relationship, a series of standard models for distributions with a right skew have been fit to the simulation results for vertical distance older than 2500 BCE (Figure \@ref(fig:results2)A) by means of maximum likelihood estimation (Table \@ref(tab:tab1)). As most of the models only accept positive values, a constant of 0.001 was added to avoid values of zero. It was attempted to both remove negative values and force these to zero before adding the constant. As the difference between these two solutions was negligible, and as the assumption here is that negative values in actuality reflect a distance of zero, the latter approach was chosen (a plot displaying the negative values and the compared models is available in the supplementary material).

@@

An example of an implementation of the outlined approach is given in Figure \@ref(fig:exampledat), where $\alpha$ = `r round(gammafit$estimate[1], 3)` and $\sigma$ = `r round(gammafit$estimate[2], 3)`. These are the parameters for the gamma distribution identified when considering all pre-Late Neolithic simulation results (Figure \@ref(fig:results2)A) and are the parameters used in all applications of the proposed method that follow below. For the numerical implementation, $D$ is here stepped through as a sequence of increments of 0.001m, which, following from the adjustment of the values for fitting the compared models, starts from 0.001m. The gamma distribution is stepped through in it's cumulative form, where the probability from the previous 0.001m step is subtracted from the probability at the current step. This probability is then divided equally across the individual calendar years in the range between the lower and the upper limit of the displacement curve at the current 0.001m step. The probability mass function that is the resulting shoreline date is the sum of performing this procedure on all possible 0.001m values of $D$, which, in practice, is down to and including $E - D = 0.001$ or when 99.999% of the gamma distribution has been stepped through.

@@

```{r exampledat, echo = FALSE, fig.cap = paste0("Shoreline dating of Hegna vest 1. The mean elevation of the site polygon is used to inform $E$ in the dating of the site. The gamma distribution in blue on the y-axis extends the full range of possible values for $E - D$ and has the parameters $\\alpha$ = ", round(gammafit$estimate[1], 3), " and $\\sigma$ = ", round(gammafit$estimate[2], 3),  " (see Figure 7A). The red envelope marks the shoreline displacement curve interpolated to the site location. The resulting shoreline date in grey is underlined with the 95\\% HDR in black."), out.width = "500px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/sdate.png"))

@@

With these considerations in mind, the results also indicate that shoreline dating has at times been applied with an exaggerated degree of precision. While the implications of a more stable RSL-change for shoreline dating are well known, this also appears to be somewhat under-appreciated in the practical implementation of the method. The results indicate that the spatial and temporal contingency of the method is better captured by the implementation suggested here, as is illustrated by the variation in the range of the 95% HDRs for the dates. In some cases the proposed method provides a relatively precise date and in others it offers little more than a *terminus post quem*. This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise dates (cf. Figure \@ref(fig:map-iso)B). However, as some of the 95% HDRs extend well beyond major chronological divisions, even into the Iron Age, it is also clear that some of these could be severely and securely constrained with only cursory reference to typology. While this would be trivial in some cases, the nature and uncertainty inherent to the method still means that this is arguably an exercise that should be explicitly performed. This also points to the possibility of drawing on other temporal data to further improve the precision of the dates that can be achieved with shoreline dating.
